---
title: "Exploratory analysis with ggplot2"
output: 
  learnr::tutorial:
    css: "css/custom.css"
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(uitk)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE, exercise=TRUE, warning=FALSE)
```

## A first visualization

For your first introduction to R, we will explore the relationship between education and income in Massachusetts.

### Familiarize yourself

There is no one best way to begin an exploratory analysis to guarantee interesting outcomes. But before one begins their EDA, they must know what their data actually contain. We will use the dataset `acs_edu` from the uitk package. Load `uitk` and the `tidyverse`. 

```{r read-data, message=FALSE, exercise.lines =2 }
library(uitk)
library(tidyverse)
```

`acs_edu` contains data demographic information about every census tract in Massachusetts. Print `acs_edu` to the console. What do you see?

```{r print-data, max.print=5}
acs_edu
```

`## # A tibble: 1,456 x 7` is printed out at the top followed by column names, their types—e.g. `<dbl>`—their respective values and, to the far left we see the numbers 1 through 10 before each row of values. 

Let us dissect `# A tibble: 1,456 x 7` a little bit more. This alone is quite informative. It tells us that the type of object we are working with is a `tibble` with 1,456 rows and 7 columns.

A tibble is a method of representing rectangular data and is very similar to a table one may create within Excel with rows an columns. When working with tibbles we try to adhere to what are called the principles of tidy data[^tidydata]. There are three key principles that we ought to keep in mind when working with rectangular data.

1. Each variable forms a column.
2. Each observation forms a row.
3. Each value represents a combination of an observation and a variable.

There can often be confusion about what should be a variable and what is to be an observation. In _Tidy Data_ Hadley Wickham write that

> "A variable contains all values that measure the same underlying attribute (like height, temperature, duration) across units. An observation contains all values measured on the same unit (like a person, or a day, or a race) across attributes."[^tidydata]

Say we have a tibble of survey respondents. In this case each row should be a respondent and each column should be a variable that is associated with that respondent. This could be something such as age, birth date, or the respondents response to a survey question. 

In the case of our `acs_edu` tibble, our unit of observation, aka row, is a census tract. Each variable measures a different characteristic of a census tract. For example, the column `med_house_income` is an estimate of the median household income of a given census tract. The other columns indicate what proportion of a population meets some criteria. 

How does one know what criteria their columns represent? This brings us to the importance of column names. Column names ought to be descriptors of their corresponding variables. This is a surprsingly difficult task! In `acs_edu` we can infer—though we should always have documentation to supplement the data—that the variables measure income, educational attainment rates, and race.

## Form a question

Once you have familiarized yourself with the data that you will be working with, you can begin to form a question that can be feasibly be explored or answered with the present data. The importance of domain expertise in EDA cannot be understated. Without an understanding of what underlying phenomena your data are measuring it will be extremely difficult to come to meaningful insights. 

My background is in sociology. Within sociology, and specifically social stratification, it is believed that more education leads to more social prestige, economic stability, and is more readily accessible by the white population. Given this background and the data available in `acs_edu`, we will explore the relationship between education and income. We will try to answer the question _what is the relationship between education and income?_ We will first look at each variable in isolation and then try and identify any relationship that may exist between the two variables.

## Building a graph

To create our visualizations we will use the package [`ggplot2`](https://ggplot2.tidyverse.org/) from the tidyverse. Before we can begin, we need to make sure that the collection of functions from ggplot2 are available to us with the `library()` function.

> Reminder: `library(pkg_name)` loads a package into your workspace and makes the functions and objects it exports available to you. 

```{r load-ggplot2}
library(ggplot2)
```

To begin building a ggplot, we use the function `ggplot()`. There are two function arguments these being `data` and the aesthetics `mapping`. The `data` is the tibble that we wish to visualize. In this case we want to visualize the data from `acs_edu`. 

We will begin constructing our first visualization with the `ggplot()` **function** using the `acs_edu` **object**.

> Reminder: Functions are characterised by the parentheses at the end of them. Functions do things whereas objects hold information. 

```{r bare-plot}
ggplot(acs_edu)
```
 
Notice that this plot is entirely empty. This is because we have not defined what it is that we want to visualize. ggplot uses what is called a grammar of graphics (this is expanded upon in depth in the _Visualizing Trends and Relationships_ chapter) which requires us to sequentially build our graphs by first defining what data and variables will be visualized and then adding layers to the plot.

The next step we need to take is to define which columns we want to visualize. These are called the _aesthetics_ and they are defined using the `aes()` function which is supplied to the `mapping` argument. The purpose of `aes()` is to tell ggplot which columns are mapped to what. The most important and fundamental of these are the `x` and `y` arguments. These refer to the x and y axes in the chart that we will begin to make.

Before we begin to analyze the relationship between `med_house_income` and `bach` (bachelor's degree attainment rate), we ought to do our due diligence of looking at the distribution of each of these first. Let us start with the `med_house_income` column. When exploring only a single variable, we want to supply that to the `x` argument of `aes()`.

```{r set-x-aes}
ggplot(acs_edu, aes(x = med_house_income))
```

Alright, we are making progress. We can see that the x axis is now filled out a bit more. The axis breaks have been labeled as has the axis itself. In order to see the data in a graphical representation, we need to determine how we want to see the data and what sort of geometry will be used to visualize it. 

To **add** geometry to our ggplot, we use the plus sign `+` which signifies that we are adding a layer on top of the basic graph. There are many ways we can visualize univariate data but the histogram has stood the test of time. To create a histogram we **add** the `geom_histogram()` layer to our existing ggplot code. 

```{r inc-hist, warning=FALSE, message=FALSE}
ggplot(acs_edu, aes(x = med_house_income)) + 
  geom_histogram()
```

> Note: To ensure that our code is legible we add each new layer on a line. R will manage the indentation for you. Code readibility is very important and you will thank yourself later for instilling good practices from the start. 

This histogram illustrates the distribution of median household income in the state of Massachusetts. The median value seems to sit somewhere around \$75k with a few outliers near \$250k as well demonstrating a right skew.

> Reminder: The skew is where there are few [observations]. 

Usually when we look at distributions of wealth they are extremely right skewed meaning there are a few people who make an outrageous amount of money. What is interesting is that this histogram is rather normally distributed almost challenging intuition. This is because the ACS does something called top-coding. Top-coding is the practice of creating a ceiling value. For example, if there is a tract has a median household income of \$1m, that will be reduced to the top-coded value—what appears to be \$250k. This creates what are called censored data. 

> **Censored data**: data "in which the value of a measurement or observation is only partially known." [^censorwiki][^censorstack] 

Let us now create a histogram of our second variable of interest, `bach`.

```{r edu-hist, warning=FALSE, message=FALSE}
ggplot(acs_edu, aes(x = bach)) + 
  geom_histogram()
```

This histogram illustrates the distribution of the bachelor degree attainment rate (the proportion of people with a bachelor's degree) across census tracts in Massachusetts. Because we did our homework ahead of time, we know that the national attainment rate in 2018 for people over 25 was ~35%[^acsedu]. Our histogram shows that within MA there is a lot of variation in the attainment rate from a low of about 0% to a high of over 60%. There is not a steep peak in the distribution which tells us that there is a fair amount of variation in the distribution.

Now that there is an intuition of the distribution and characteristics of both `med_house_income` and `bach`, we can begin to try and answer the question _what is the effect of education on median household income?_ The phrasing of our question will determine how we visualize our data. 

When stating research questions we often phrase it as _what is the effect of x **on** y_? In this formulation we are determining that `bach`, our independent variable, will be plotted on the x axis and `med_house_income` will be plotted on the y axis. To visualize this bivariate relationship we will create a scatter plot. 

> This structure and phrasing is useful for continuity in verbal communication, graphical representation, and hypothesis testing. 

We can visualize this relationship by adding additional mapped aesthetics. In this case, we will map both the x and y arguments of the `aes()` function. Rather than adding histogram layer, we will need to create a scatter plot. Scatter plots are created by plotting points for each (x, y) pair. To get such an effect we will use the `geom_point()` layer. 

```{r inc-edu-scatter}
ggplot(acs_edu, aes(x = bach, y = med_house_income)) +
  geom_point()
```

The above scatter plot provides a lot of information. We see that there is a positive linear trend—that is that when the `bach` value increases so does the `med_house_income` variable. When looking at a scatter plot we are looking to see if there is a consistent pattern that can be sussed out. 

In this scatterplot we can see that there is a linear pattern. When the points on the scatter plot are closer to eachother and demonstrate less spread, that means there is a stronger relationship between the two variables. Imagine if we drew a line through the middle of the points, we would want each point to be as close to that line as possible. The further the point is away from that line, the more variation there is. In these cases we often create linear regression models to estimate the relationship.

Using ggplot, we can plot the estimated linear regression line on top of our scatterplot. This is done with a `geom_smooth()` layer. By default, `geom_smooth()` does not plot the linear relationship. To do that, we need to specify what kind of smoothing we would like. To plot the estimated linear model, we set `method = "lm"`.  

```{r scatter-smooth}
ggplot(acs_edu, aes(x = bach, y = med_house_income)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Wonderful! To finish up this graphic, we should add informative labels. Labels live in their own layer which is created with `labs()`. Each argument maps to an aesthetic—e.g. `x` and `y`. By default ggplot uses the column names for axis labels, but these labels are usually uninformative.

Let's give the plot a title and better labels for its axes. We will set the following arguments to `labs()`

  - `x = "% of population with a Bachelor's Degree"`
  - `y = "Median Household Income"`
  - `title = "Relationship between Education and Income"`

```{r plot-labelled}
ggplot(acs_edu, aes(x = bach, y = med_house_income)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "% of population with a Bachelor's Degree",
       y = "Median Household Income",
       title = "Relationship between Education and Income")
```

> Note that each argument is placed on a new line. Again, this is to improve readability.

What can we determine from this graph? Take a few minutes and write down what you see and what you can infer from that. 

> Consider asking these questions:
> 
- Is there a relationship between our variables?
- Is the relationship linear?
- Which direction does the trend go?
- How spread out are the value pairs?
- Are there any outliers? 

This chart ilicits further lines of inquiry. For example, in the sociological literature there is a well documented achievement gap. The achievement gap can be traced along racial lines—though it is not inherently _caused_ by race but rather intertwined with it. Can this be seen in the tracts of Massachusetts? 

We can visualize a third variable on our chart by mapping another `aes`thetic. When we add a third variable to the visualization we are generally trying to illustrate group membership or size / magnitude of the third variable. Due to the large number of points on our chart already, we may benefit more from mapping color rather than size—imagine 1,000 points overlapping even more than they already do. 

We can map the proportion of the population that is white to the color of our points. We do this by setting the `color` aesthetic to `white`. While we're at it, let us include a subtitle which is informative to the viewer.

```{r colored-plot}
ggplot(acs_edu, aes(x = bach, y = med_house_income, color = white)) +
  geom_point(alpha = .4) +
  geom_smooth(method = "lm") + 
  labs(x = "% of population with a Bachelor's Degree",
       y = "Median Household Income",
       title = "Relationship between Education and Income",
       subtitle = "Colored by whiteness") 
```

What can we conclude now? Does the addition of the third variable increase or decrease the utility of our scatter plot? Does the trend seem to mediated by race? I'll leave those questions to you to answer.

You've now completed your first visual analysis. You've learned how to create publication ready histograms and scatter plots using ggplot2. This is no small feat! 

This chapter provided you with data that was used in our visualization exercises. You're going to want to be able to visualize and analyze your own data. The next chapter introduces you reading data and some of the most common file formats you may encounter.

[^12]: https://www.census.gov/history/www/through_the_decades/index_of_questions/1790_1.html
[^13]: https://www2.census.gov/programs-surveys/acs/methodology/design_and_methodology/acs_design_methodology_report_2014.pdf
[^14]: https://www.census.gov/programs-surveys/acs/methodology.html
[^15]: https://www.nytimes.com/2012/05/20/sunday-review/the-debate-over-the-american-community-survey.html
[^16]: https://www.census.gov/content/dam/Census/library/publications/2018/acs/acs_general_handbook_2018_ch03.pdf
[^tidydata]: https://vita.had.co.nz/papers/tidy-data.pdf
[^style]: https://style.tidyverse.org/index.html
[^censorwiki]: https://en.wikipedia.org/wiki/Censoring_(statistics)
[^censorstack]: https://stats.stackexchange.com/questions/49443/how-to-model-this-odd-shaped-distribution-almost-a-reverse-j
[^acsedu]: https://www.census.gov/data/tables/2018/demo/education-attainment/cps-detailed-tables.html
[^nist]: NIST Handbook. https://www.itl.nist.gov/div898/handbook/eda/section1/eda11.htm.
