---
title: "Reading Data"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, exercise=TRUE, warning=FALSE)
requests <- readr::read_csv("https://raw.githubusercontent.com/JosiahParry/uitk/master/data/311-sample.csv")
```


## Reading Data 

First, read [Chapter 8](https://ui.josiahparry.com/reading-data.html) of Urban Informatics Toolkit.

To read data we will use the package readr from the tidyverse. 

We provide the path to the data. Most often this will be a file path on your computer like `data/`. Or, it can be a URL. In this example we will be using a file hosted online. It is at `https://raw.githubusercontent.com/JosiahParry/uitk/master/data/311-sample.csv`. Go ahead and give that a look to see the raw data. Supply that as a character string (surround by quotes like `"words and stuff"`). When things are wrapped in quotations R treats them like words and not objects we've created. 


## Reading CSVs

1. Read in the CSV at `https://raw.githubusercontent.com/JosiahParry/uitk/master/data/311-sample.csv` and save it to the object requests. 
2. Print out the resultant object

```{r reading-csv, message=FALSE}
library(tidyverse)

requests <- read_csv("https://raw.githubusercontent.com/JosiahParry/uitk/master/data/311-sample.csv")

```


## Reading Excel Files

```{r}
library(readxl)

readxl::read_excel("https://raw.githubusercontent.com/JosiahParry/uitk/master/data/311-sample.xlsx")
```


## Tab delimited data



## Looking at your data

- `glimpse()` will preview all of the columns in your data frame.

```{r glimpse}
glimpse(requests)
```

- `summary()`

```{r summary}
summary(requests)
```

### {skimr}

skimr is a package that provides a nice visual summary of your data. It summarizes each type of data differently. 

1. Install the `skimr` package
2. Use the function `skim()` on your requests data frame.

Since you only need one function it makes more sense to call the function directly without having to load the entire package. You would do this like `skimr::skim(dataframe)`.

```{r skimr}
skimr::skim(requests)
```

### The shape of your data

Sometimes you will need to know the dimensions of your data set in a programmatic way. There are, of course, functions in R to help you do this. The function `dim()` will return a **vector** of length two. Containing first, the number of rows, and second, the number of columns. 

```{r dims}
dim(requests)
```

I think we ought to be a bit more explicit, though, and prevent any room for confusion. To retrieve the number of rows and column use the functions `nrow()` and `ncol()` respectively. These only return on value—no need to remember if the first or second value is the one you want! 

```{r nrow-ncol}
nrow(requests)

ncol(requests)
```


csv
excel 
tsv


